---
title: "Characterizing Sequence-Function Relationships in Chimeric DcuS/EnvZ Histidine Kinases"
author: 'Authors: <font color="blue">Luca B. Lippert, </font color="black">Samuel R. Hinton, Andrew Holston, Karl J. Romanowicz, Calin Plesa'
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
      smooth_scroll: yes
  pdf_document:
    toc: yes
    toc_depth: '5'
---

<font color="black"> **R Notebook:** <font color="green">Provides reproducible analysis for **Association of mutant sequence, barcode, and inferred fluorescence phenotype** in the following manuscript: <font color="black">

**Citation:** Lippert LB, Hinton SR, Holston A, Romanowicz KJ, Plesa C. Characterizing Sequence-Function Relationships in Chimeric DcuS/EnvZ Histidine Kinases. *In Prep.* 2026.

**GitHub Repository:** <https://github.com/PlesaLab/DcuSEnvZ>

# Experiment

<font color="black">This pipeline processes</font> <font color="red">*barcode counts sequenced from twelve fluroescence-activated cell sorting (FACS) samples of the synTCS-MutLib strain in the aspartate condition*. <font color="black">Sequence data was generated using the Illumina NextSeq platform using paired-end sequencing read amplicons. Raw sequencing data was pre-processed on a high-performance computer using the Makefile script available in the project GitHub repository. Here, pre-processed barcode-count files for all twelve aspartate samples are merged, adjusted abundances are computed using the protocol from Biswas et al., (2021), and a median activation score- here, "median bin"- is calculated for each observed barcode. This analysis is replicated for the No Ligand and Fumarate samples.

```{css, echo=FALSE}
.badCode {
background-color: lightpink;
font-weight: bold;
}

.goodCode {
background-color: lightgreen;
font-weight: bold;
}

table {
  margin: auto;
  border-top: 1px solid #666;
  border-bottom: 1px solid #666;
}
table thead th { border-bottom: 1px solid #ddd; }
th, td { padding: 5px; }
thead, tfoot, tr:nth-child(even) { background: #eee; }
```

```{r setup, include=FALSE, echo=FALSE}
# Set global options for notebook
knitr::opts_chunk$set(
  echo = TRUE,
  warning = TRUE,
  message = TRUE,
  results = "markup",
  collapse = FALSE
)
# Getting the path of your current open file and set as wd
current_path = rstudioapi::getActiveDocumentContext()$path 
setwd(dirname(current_path))
print(getwd())
```

# Packages

The following R packages must be installed prior to loading into the R session. See the **Reproducibility** tab for a complete list of packages and their versions used in this workflow.

```{r message=FALSE, results='hide', warning=FALSE}
# Make a vector of required packages
required.packages <- c("devtools", "knitr", "patchwork", "tidyverse", "ggplot2", "dplyr", "tidyr", "magrittr", "stringr", "seqinr")

# Load required packages
lapply(required.packages, library, character.only = TRUE)
```

```{r include=FALSE}
# set.seed is used to fix the random number generation to make the results repeatable
set.seed(123)
```

# Barcode Counts to Median Bin Pipeline

<font color="blue">**This section is based on the R file: "Counts_to_Median_Bin_Aspartate.R".**<font color="black"> It describes how to load all of the pre-existing barcode data necessary for downstream analysis. The end result is a .CSV file containing the total set of observed barcodes, their associated nucleotide and amino acid sequences, activation ("median bin") scores, the lower and upper indices of their activation bins.

## Read in data

```{r}
# Function to load barcode/BC reads
read_collapsed_file <- function(filename, sam_name) {
  df <- read.table(file=filename, sep="\t", header=FALSE)
  colnames(df) <- c("BC", paste0(sam_name, "reads"), "collapsedBCs")
  return(df)
}

A1_bc <- read_collapsed_file("./Final_BC/A1_S25_collapse_d1.tsv", "A1")
A2_bc <- read_collapsed_file("./Final_BC/A2_S26_collapse_d1.tsv", "A2")
A3_bc <- read_collapsed_file("./Final_BC/A3_S27_collapse_d1.tsv", "A3")
A4_bc <- read_collapsed_file("./Final_BC/A4_S28_collapse_d1.tsv", "A4")
A5_bc <- read_collapsed_file("./Final_BC/A5_S29_collapse_d1.tsv", "A5")
A6_bc <- read_collapsed_file("./Final_BC/A6_S30_collapse_d1.tsv", "A6")
A7_bc <- read_collapsed_file("./Final_BC/A7_S31_collapse_d1.tsv", "A7")
A8_bc <- read_collapsed_file("./Final_BC/A8_S32_collapse_d1.tsv", "A8")
A9_bc <- read_collapsed_file("./Final_BC/A9_S33_collapse_d1.tsv", "A9")
A10_bc <- read_collapsed_file("./Final_BC/A10_S34_collapse_d1.tsv", "A10")
A11_bc <- read_collapsed_file("./Final_BC/A11_S35_collapse_d1.tsv", "A11")
A12_bc <- read_collapsed_file("./Final_BC/A12_S36_collapse_d1.tsv", "A12")
```

## Combine and clean the total dataset for aspartate barcodes

Put all BCs into one dataframe for each condition

```{r}
A_allBC <- A1_bc %>%
  select(BC) %>%
  rbind(., A2_bc %>% 
          select(BC)) %>%
  rbind(., A3_bc %>% 
          select(BC)) %>%
  rbind(., A4_bc %>% 
          select(BC)) %>%
  rbind(., A5_bc %>% 
          select(BC)) %>%
  rbind(., A6_bc %>% 
          select(BC)) %>%
  rbind(., A7_bc %>% 
          select(BC)) %>%
  rbind(., A8_bc %>% 
          select(BC)) %>%
  rbind(., A9_bc %>% 
          select(BC)) %>%
  rbind(., A10_bc %>% 
          select(BC)) %>%
  rbind(., A11_bc %>% 
          select(BC)) %>%
  rbind(., A12_bc %>% 
          select(BC)) %>%
  distinct()
```

Add counts for barcodes

```{r}
A_allBC <- left_join(A_allBC, A1_bc %>% select(-collapsedBCs), by="BC")
A_allBC <- left_join(A_allBC, A2_bc %>% select(-collapsedBCs), by="BC")
A_allBC <- left_join(A_allBC, A3_bc %>% select(-collapsedBCs), by="BC") 
A_allBC <- left_join(A_allBC, A4_bc %>% select(-collapsedBCs), by="BC") 
A_allBC <- left_join(A_allBC, A5_bc %>% select(-collapsedBCs), by="BC")
A_allBC <- left_join(A_allBC, A6_bc %>% select(-collapsedBCs), by="BC") 
A_allBC <- left_join(A_allBC, A7_bc %>% select(-collapsedBCs), by="BC") 
A_allBC <- left_join(A_allBC, A8_bc %>% select(-collapsedBCs), by="BC") 
A_allBC <- left_join(A_allBC, A9_bc %>% select(-collapsedBCs), by="BC") 
A_allBC <- left_join(A_allBC, A10_bc %>% select(-collapsedBCs), by="BC") 
A_allBC <- left_join(A_allBC, A11_bc %>% select(-collapsedBCs), by="BC") 
A_allBC <- left_join(A_allBC, A12_bc %>% select(-collapsedBCs), by="BC")
```

Filter all barcodes by length.

```{r}
A_allBC <- A_allBC %>%
  filter(str_length(BC) >= 24)
```

If not in a set, force to 0

```{r}
A_allBC$A1reads[is.na(A_allBC$A1reads)] <- 0
A_allBC$A2reads[is.na(A_allBC$A2reads)] <- 0
A_allBC$A3reads[is.na(A_allBC$A3reads)] <- 0
A_allBC$A4reads[is.na(A_allBC$A4reads)] <- 0
A_allBC$A5reads[is.na(A_allBC$A5reads)] <- 0
A_allBC$A6reads[is.na(A_allBC$A6reads)] <- 0
A_allBC$A7reads[is.na(A_allBC$A7reads)] <- 0
A_allBC$A8reads[is.na(A_allBC$A8reads)] <- 0
A_allBC$A9reads[is.na(A_allBC$A9reads)] <- 0
A_allBC$A10reads[is.na(A_allBC$A10reads)] <- 0
A_allBC$A11reads[is.na(A_allBC$A11reads)] <- 0
A_allBC$A12reads[is.na(A_allBC$A12reads)] <- 0

rm(A1_bc,A2_bc,A3_bc,A4_bc,A5_bc,A6_bc,A7_bc,A8_bc,A9_bc,A10_bc,A11_bc,A12_bc)
```

## Normalize barcode counts to the actual fraction of the population they represent

Without doing this, the bins are weighted equally (pct. total pop = 8.3333%), when in actuality, the population was not perfectly distributed across bins. Population values were calculated by dividing the number of events recorded per bin by total population in a Python script, synTCS-MutLib_FACS_Bin_Population_Fractions.ipynb.

```{r}
A_bin1_NormVal = 8.3333333/3.66423061
A_bin2_NormVal = 8.3333333/8.85099472
A_bin3_NormVal = 8.3333333/13.07348762
A_bin4_NormVal = 8.3333333/9.41940723
A_bin5_NormVal = 8.3333333/11.87576127
A_bin6_NormVal = 8.3333333/11.81485993
A_bin7_NormVal = 8.3333333/9.59196102
A_bin8_NormVal = 8.3333333/7.63296793
A_bin9_NormVal = 8.3333333/8.51603735
A_bin10_NormVal = 8.3333333/8.46528624
A_bin11_NormVal = 8.3333333/5.21721478
A_bin12_NormVal = 8.3333333/1.87779131

A_allBC <- A_allBC %>%
  mutate(A1reads_corrected = A1reads / A_bin1_NormVal,
         A2reads_corrected = A2reads / A_bin2_NormVal,
         A3reads_corrected = A3reads / A_bin3_NormVal,
         A4reads_corrected = A4reads / A_bin4_NormVal,
         A5reads_corrected = A5reads / A_bin5_NormVal,
         A6reads_corrected = A6reads / A_bin6_NormVal,
         A7reads_corrected = A7reads / A_bin7_NormVal,
         A8reads_corrected = A8reads / A_bin8_NormVal,
         A9reads_corrected = A9reads / A_bin9_NormVal,
         A10reads_corrected = A10reads / A_bin10_NormVal,
         A11reads_corrected = A11reads / A_bin11_NormVal,
         A12reads_corrected = A12reads / A_bin12_NormVal
  ) 
```

## Compute a relative abundances matrix

Biswas et al., 2021: Compute a relative abundance table, R, by dividing the columns of C by their sums. The columns of R sum to 1.

```{r}
A1_total <- sum(A_allBC$A1reads_corrected)
A2_total <- sum(A_allBC$A2reads_corrected)
A3_total <- sum(A_allBC$A3reads_corrected)
A4_total <- sum(A_allBC$A4reads_corrected)
A5_total <- sum(A_allBC$A5reads_corrected)
A6_total <- sum(A_allBC$A6reads_corrected)
A7_total <- sum(A_allBC$A7reads_corrected)
A8_total <- sum(A_allBC$A8reads_corrected)
A9_total <- sum(A_allBC$A9reads_corrected)
A10_total <- sum(A_allBC$A10reads_corrected)
A11_total <- sum(A_allBC$A11reads_corrected)
A12_total <- sum(A_allBC$A12reads_corrected)

A_allBC_R <- A_allBC %>%
  mutate(A1_norm=A1reads_corrected/A1_total,
         A2_norm=A2reads_corrected/A2_total,
         A3_norm=A3reads_corrected/A3_total,
         A4_norm=A4reads_corrected/A4_total,
         A5_norm=A5reads_corrected/A5_total,
         A6_norm=A6reads_corrected/A6_total,
         A7_norm=A7reads_corrected/A7_total,
         A8_norm=A8reads_corrected/A8_total,
         A9_norm=A9reads_corrected/A9_total,
         A10_norm=A10reads_corrected/A10_total,
         A11_norm=A11reads_corrected/A11_total,
         A12_norm=A12reads_corrected/A12_total) %>%
  select(BC, A1_norm, A2_norm, A3_norm, A4_norm, A5_norm, A6_norm, A7_norm, A8_norm, A9_norm, A10_norm, A11_norm, A12_norm) %>%
  dplyr::rename(barcode=BC)

# Check, sum off all values in each column should equal 1
sum(A_allBC_R$A1_norm)
```

Calculate the total reads for each barcode across all bins

```{r}
A_allBC_total_counts <- A_allBC %>% 
  mutate(BC_SorTotReads = A1reads_corrected + A2reads_corrected + A3reads_corrected + A4reads_corrected + A5reads_corrected + A6reads_corrected + A7reads_corrected + A8reads_corrected + A9reads_corrected + A10reads_corrected + A11reads_corrected + A12reads_corrected) %>% 
  select(BC, BC_SorTotReads) %>% 
  dplyr::rename(barcode=BC)
```

## Merge barcode-relative abundances with nucleotide sequences

Read in barcode-nucleotide sequence mapping file.

```{r}
consensus_gene <- read.csv(file="./input_files/consensus_gene.csv",head=TRUE,sep=",")
consensus_gene %>% select(description) %>% distinct() %>% nrow()  # 951065 unique barcodes

consensus_gene2 <- consensus_gene %>%
  select(description,sequence) %>%
  dplyr::rename(barcode=description,NTseq=sequence)

# convert to strings
str(consensus_gene2)
```

Merge from all bins and create a master list of unique variants and BCs

```{r}
tgood_A <- right_join(consensus_gene2, A_allBC_R, join_by(barcode)) %>%
  mutate(NTlen=nchar(as.character(NTseq)))

sum(is.na(tgood_A$NTseq)) #471038 barcodes in A_allBC_R don't have a nucleotide sequence associated with them

# Normalization check, sum of all values in each column should still be 1
sum(tgood_A$A4_norm)
```

Mapping check: how many barcodes have more than 1 variant?

```{r}
consensus_gene_sum <- consensus_gene2 %>%
  group_by(barcode) %>%
  summarise(count=n())

consensus_gene_sum %>%
  filter(count>1) %>%
  nrow(.) 
```

Make a list of BCs with only 1 variant and filter the dataset to keep only barcodes which have been mapped.

```{r}
bcgood_A <- consensus_gene_sum %>%
  filter(count==1) %>%
  select(-count)

# Filter to only keep barcodes which appear
A_allBC_R_NTfilter <- tgood_A %>%
  semi_join(bcgood_A,by="barcode") %>% 
  left_join(A_allBC_total_counts,by="barcode")
```

## Compute total normalized barcode counts for each bin

```{r}
A_allBC_R_NTfilter_totals <- A_allBC_R_NTfilter %>%
  group_by(barcode) %>%
  summarise(NTseq=NTseq,
            A1_t=sum(A1_norm),
            A2_t=sum(A2_norm),
            A3_t=sum(A3_norm),
            A4_t=sum(A4_norm),
            A5_t=sum(A5_norm),
            A6_t=sum(A6_norm),
            A7_t=sum(A7_norm),
            A8_t=sum(A8_norm),
            A9_t=sum(A9_norm),
            A10_t=sum(A10_norm),
            A11_t=sum(A11_norm),
            A12_t=sum(A12_norm),
            BC_SorTotReads=BC_SorTotReads)
```

## Merge barcode-relative abundances with amino acid sequences

Load file of synTCS-MutLib variants - amino acid sequences - and filter to only keep translated sequences

```{r}
consensus_prot <- read.csv(file="./input_files/consensus_prot_with_PreSortBC.csv",head=TRUE,sep=",")

A_allBC_R_AAfilter <- A_allBC_R_NTfilter_totals %>%
  left_join(consensus_prot %>% dplyr::rename(barcode=BC),by="barcode")

# Some sequences have mutations which place a stop codon at the beginning and some barcodes were not mapped to amino acid sequences; filter these out.
A_allBC_R_AAfilter <- A_allBC_R_AAfilter %>%
  filter(!is.na(seq))
```

Replace read counts of NA and 0 with an arbitrary value of 0.1 for presort1 and presort2 libraries

```{r}
A_allBC_R_AAfilter$presort1reads[is.na(A_allBC_R_AAfilter$presort1reads)] <- 0.1
A_allBC_R_AAfilter$presort1reads[A_allBC_R_AAfilter$presort1reads == 0] <- 0.1
A_allBC_R_AAfilter$presort2reads[is.na(A_allBC_R_AAfilter$presort2reads)] <- 0.1
A_allBC_R_AAfilter$presort2reads[A_allBC_R_AAfilter$presort2reads == 0] <- 0.1
```

## Calculate a fold-change matrix

Biswas et al., 2021: Divide each column of R element-wise by the input relative abundance vector (relative abundance of variants in the library before flow cytometry) to obtain a fold-change table, F.

```{r}
A_allBC_F <- A_allBC_R_AAfilter %>%
  mutate(A1_fc=A1_t/presort2_norm,
         A2_fc=A2_t/presort2_norm,
         A3_fc=A3_t/presort2_norm,
         A4_fc=A4_t/presort2_norm,
         A5_fc=A5_t/presort2_norm,
         A6_fc=A6_t/presort2_norm,
         A7_fc=A7_t/presort2_norm,
         A8_fc=A8_t/presort2_norm,
         A9_fc=A9_t/presort2_norm,
         A10_fc=A10_t/presort2_norm,
         A11_fc=A11_t/presort2_norm,
         A12_fc=A12_t/presort2_norm)
```

## Calculate an adjusted abundances table

Biswas et al., 2021: Divide each row of F by its sum to obtain a table of adjusted abundances, A. Each row of A sums to 1.

```{r}
A_allBC_A <- A_allBC_F %>%
  mutate(rowsum=A1_fc+A2_fc+A3_fc+A4_fc+A5_fc+A6_fc+A7_fc+A8_fc+A9_fc+A10_fc+A11_fc+A12_fc) %>%
  mutate(A1=A1_fc/rowsum,
         A2=A2_fc/rowsum,
         A3=A3_fc/rowsum,
         A4=A4_fc/rowsum,
         A5=A5_fc/rowsum,
         A6=A6_fc/rowsum,
         A7=A7_fc/rowsum,
         A8=A8_fc/rowsum,
         A9=A9_fc/rowsum,
         A10=A10_fc/rowsum,
         A11=A11_fc/rowsum,
         A12=A12_fc/rowsum)
```

## Determine median bin for barcodes via cumulative sum across adjusted abundances

```{r}
# create functions to compute upper and lower index
maxcs = function(x, output){
  return(max(which(c(x[1],x[2],x[3],x[4],x[5],x[6],x[7],x[8],x[9],x[10],x[11],x[12]) < 0.5)))
}
mincs = function(x, output){
  return(min(which(c(x[1],x[2],x[3],x[4],x[5],x[6],x[7],x[8],x[9],x[10],x[11],x[12]) >= 0.5)))
}

# Compute cumulative sum across adjusted barcode abundances for all bins to estimate median bin
A_allBC_CS <- A_allBC_A %>%
  rowwise() %>%
  mutate(
    cumulative_p = list(cumsum(c(A1,A2,A3,A4,A5,A6,A7,A8,A9,A10,A11,A12))),
    lower_index = max(which(cumsum(c(A1,A2,A3,A4,A5,A6,A7,A8,A9,A10,A11,A12)) < 0.5)),
    upper_index = min(which(cumsum(c(A1,A2,A3,A4,A5,A6,A7,A8,A9,A10,A11,A12)) >= 0.5)),
    median = ifelse(
      is.infinite(lower_index), 
      1, 
      lower_index + (0.5 - unlist(cumulative_p)[lower_index]) / 
        (unlist(cumulative_p)[upper_index] - unlist(cumulative_p)[lower_index])
    )
  )
```

## Export the data as a CSV file

```{r}
A_allBC_CS_toCSV <- A_allBC_CS %>%
  select(barcode, NTseq, seq, lower_index, upper_index, median, BC_SorTotReads, presort1reads, presort2reads)

write.csv(A_allBC_CS_toCSV,
          "./output_files/DcuS_Aspartate_bin_distribution-byBC.csv", row.names = FALSE)

```

# Reproducibility

The session information is provided for full reproducibility.

```{r}
devtools::session_info()
```

# References

Biswas, S.; Khimulya, G.; Alley, E. C.; Esvelt, K. M.; Church, G. M. Low-N Protein Engineering with Data-Efficient Deep Learning. Nat. Methods 2021, 18 (4), 389–396. <https://doi.org/10.1038/s41592-021-01100-y>.
