---
title: "Characterizing Sequence-Function Relationships in Chimeric DcuS/EnvZ Histidine Kinases"
author: 'Authors: <font color="blue">Luca B. Lippert, </font color="black">Samuel R. Hinton, Andrew Holston, Karl J. Romanowicz, Calin Plesa'
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
      smooth_scroll: yes
  pdf_document:
    toc: yes
    toc_depth: '5'
---
**R Notebook:** <font color="green">Provides reproducible analysis for **Association of mutant sequence, barcode, and inferred fluorescence phenotype** in the following manuscript:  <font color="black">
  
  **Citation:** Lippert LB, Hinton SR, Holston A, Romanowicz KJ, Plesa C. Characterizing Sequence-Function Relationships in Chimeric DcuS/EnvZ Histidine Kinases. *In Prep.* 2026.  
  
  **GitHub Repository:** [https://github.com/PlesaLab/DcuSEnvZ](https://github.com/PlesaLab/DcuSEnvZ)
  
# Experiment


<font color="black">This pipeline processes</font> <font color="red">*barcode counts sequenced from twelve fluroescence-activated cell sorting (FACS) samples of the synTCS-MutLib strain in the no ligand condition*. <font color="black">Sequence data was generated using the Illumina NextSeq platform using paired-end sequencing read amplicons. Raw sequencing data was pre-processed on a high-performance computer using the Makefile script available in the project GitHub repository. Here, pre-processed barcode-count files for all twelve no ligand samples are merged, adjusted abundances are computed using the protocol from Biswas et al., (2021), and a median activation score- here, "median bin"- is calculated for each observed barcode. This analysis is replicated for the Fumarate and Aspartate samples. 

```{css, echo=FALSE}
.badCode {
background-color: lightpink;
font-weight: bold;
}

.goodCode {
background-color: lightgreen;
font-weight: bold;
}

table {
  margin: auto;
  border-top: 1px solid #666;
  border-bottom: 1px solid #666;
}
table thead th { border-bottom: 1px solid #ddd; }
th, td { padding: 5px; }
thead, tfoot, tr:nth-child(even) { background: #eee; }
```

```{r setup, include=FALSE, echo=FALSE}
# Set global options for notebook
knitr::opts_chunk$set(
  echo = TRUE,
  warning = TRUE,
  message = TRUE,
  results = "markup",
  collapse = FALSE
)
# Getting the path of your current open file and set as wd
current_path = rstudioapi::getActiveDocumentContext()$path 
setwd(dirname(current_path))
print(getwd())
```

# Packages
The following R packages must be installed prior to loading into the R session. See the **Reproducibility** tab for a complete list of packages and their versions used in this workflow.
```{r message=FALSE, results='hide', warning=FALSE}
# Make a vector of required packages
required.packages <- c("devtools", "knitr", "patchwork", "tidyverse", "ggplot2", "dplyr", "tidyr", "magrittr", "stringr", "seqinr")

# Load required packages
lapply(required.packages, library, character.only = TRUE)
```

```{r include=FALSE}
# set.seed is used to fix the random number generation to make the results repeatable
set.seed(123)
```

# Barcode Counts to Median Bin Pipeline
<font color="blue">**This section is based on the R file: "Counts_to_Median_Bin_NoLigand.R".**<font color="black"> It describes how to load all of the pre-existing barcode data necessary for downstream analysis. The end result is a .CSV file containing the total set of observed barcodes, their associated nucleotide and amino acid sequences, activation ("median bin") scores, the lower and upper indices of their activation bins.  

## Read in data

```{r}
# Function to load barcode/BC reads
read_collapsed_file <- function(filename, sam_name) {
  df <- read.table(file=filename, sep="\t", header=FALSE)
  colnames(df) <- c("BC", paste0(sam_name, "reads"), "collapsedBCs")
  return(df)
}

NL1_bc <- read_collapsed_file("./Final_BC/NL1_S1_collapse_d1.tsv", "NL1")
NL2_bc <- read_collapsed_file("./Final_BC/NL2_S2_collapse_d1.tsv", "NL2")
NL3_bc <- read_collapsed_file("./Final_BC/NL3_S3_collapse_d1.tsv", "NL3")
NL4_bc <- read_collapsed_file("./Final_BC/NL4_S4_collapse_d1.tsv", "NL4")
NL5_bc <- read_collapsed_file("./Final_BC/NL5_S5_collapse_d1.tsv", "NL5")
NL6_bc <- read_collapsed_file("./Final_BC/NL6_S6_collapse_d1.tsv", "NL6")
NL7_bc <- read_collapsed_file("./Final_BC/NL7_S7_collapse_d1.tsv", "NL7")
NL8_bc <- read_collapsed_file("./Final_BC/NL8_S8_collapse_d1.tsv", "NL8")
NL9_bc <- read_collapsed_file("./Final_BC/NL9_S9_collapse_d1.tsv", "NL9")
NL10_bc <- read_collapsed_file("./Final_BC/NL10_S10_collapse_d1.tsv", "NL10")
NL11_bc <- read_collapsed_file("./Final_BC/NL11_S11_collapse_d1.tsv", "NL11")
NL12_bc <- read_collapsed_file("./Final_BC/NL12_S12_collapse_d1.tsv", "NL12")

```
## Combine and clean the total dataset for no ligand barcodes  

Put all BCs into one dataframe for each condition  
```{r}
NL_allBC <- NL1_bc %>%
  select(BC) %>%
  rbind(., NL2_bc %>% 
          select(BC)) %>%
  rbind(., NL3_bc %>% 
          select(BC)) %>%
  rbind(., NL4_bc %>% 
          select(BC)) %>%
  rbind(., NL5_bc %>% 
          select(BC)) %>%
  rbind(., NL6_bc %>% 
          select(BC)) %>%
  rbind(., NL7_bc %>% 
          select(BC)) %>%
  rbind(., NL8_bc %>% 
          select(BC)) %>%
  rbind(., NL9_bc %>% 
          select(BC)) %>%
  rbind(., NL10_bc %>% 
          select(BC)) %>%
  rbind(., NL11_bc %>% 
          select(BC)) %>%
  rbind(., NL12_bc %>% 
          select(BC)) %>%
  distinct()
```
Add counts for barcodes  
```{r}
NL_allBC <- left_join(NL_allBC, NL1_bc %>% select(-collapsedBCs), by="BC")
NL_allBC <- left_join(NL_allBC, NL2_bc %>% select(-collapsedBCs), by="BC")
NL_allBC <- left_join(NL_allBC, NL3_bc %>% select(-collapsedBCs), by="BC") 
NL_allBC <- left_join(NL_allBC, NL4_bc %>% select(-collapsedBCs), by="BC") 
NL_allBC <- left_join(NL_allBC, NL5_bc %>% select(-collapsedBCs), by="BC")
NL_allBC <- left_join(NL_allBC, NL6_bc %>% select(-collapsedBCs), by="BC") 
NL_allBC <- left_join(NL_allBC, NL7_bc %>% select(-collapsedBCs), by="BC") 
NL_allBC <- left_join(NL_allBC, NL8_bc %>% select(-collapsedBCs), by="BC") 
NL_allBC <- left_join(NL_allBC, NL9_bc %>% select(-collapsedBCs), by="BC") 
NL_allBC <- left_join(NL_allBC, NL10_bc %>% select(-collapsedBCs), by="BC") 
NL_allBC <- left_join(NL_allBC, NL11_bc %>% select(-collapsedBCs), by="BC") 
NL_allBC <- left_join(NL_allBC, NL12_bc %>% select(-collapsedBCs), by="BC")
```
Filter all barcodes by length. 
```{r}
NL_allBC <- NL_allBC %>%
  filter(str_length(BC) >= 24)
```
If not in a set, force to 0  
```{r}
NL_allBC$NL1reads[is.na(NL_allBC$NL1reads)] <- 0
NL_allBC$NL2reads[is.na(NL_allBC$NL2reads)] <- 0
NL_allBC$NL3reads[is.na(NL_allBC$NL3reads)] <- 0
NL_allBC$NL4reads[is.na(NL_allBC$NL4reads)] <- 0
NL_allBC$NL5reads[is.na(NL_allBC$NL5reads)] <- 0
NL_allBC$NL6reads[is.na(NL_allBC$NL6reads)] <- 0
NL_allBC$NL7reads[is.na(NL_allBC$NL7reads)] <- 0
NL_allBC$NL8reads[is.na(NL_allBC$NL8reads)] <- 0
NL_allBC$NL9reads[is.na(NL_allBC$NL9reads)] <- 0
NL_allBC$NL10reads[is.na(NL_allBC$NL10reads)] <- 0
NL_allBC$NL11reads[is.na(NL_allBC$NL11reads)] <- 0
NL_allBC$NL12reads[is.na(NL_allBC$NL12reads)] <- 0

rm(NL1_bc,NL2_bc,NL3_bc,NL4_bc,NL5_bc,NL6_bc,NL7_bc,NL8_bc,NL9_bc,NL10_bc,NL11_bc,NL12_bc)
```
## Normalize barcode counts to the actual fraction of the population they represent  
Without doing this, the bins are weighted equally (pct. total pop = 8.3333%), when in actuality, the population was not perfectly distributed across bins. Population values were calculated by dividing the number of events recorded per bin by total population in a Python script, synTCS-MutLib_FACS_Bin_Population_Fractions.ipynb.  

```{r}
NL_bin1_NormVal = 8.3333333/9.41426834
NL_bin2_NormVal = 8.3333333/11.69436563
NL_bin3_NormVal = 8.3333333/9.03931901
NL_bin4_NormVal = 8.3333333/8.17794893
NL_bin5_NormVal = 8.3333333/10.11349818
NL_bin6_NormVal = 8.3333333/11.01540332
NL_bin7_NormVal = 8.3333333/9.92095663
NL_bin8_NormVal = 8.3333333/8.08674503
NL_bin9_NormVal = 8.3333333/8.84677746
NL_bin10_NormVal = 8.3333333/7.78273206
NL_bin11_NormVal = 8.3333333/4.73246859
NL_bin12_NormVal = 8.3333333/1.17551682

NL_allBC <- NL_allBC %>%
  mutate(NL1reads_corrected = NL1reads / NL_bin1_NormVal,
         NL2reads_corrected = NL2reads / NL_bin2_NormVal,
         NL3reads_corrected = NL3reads / NL_bin3_NormVal,
         NL4reads_corrected = NL4reads / NL_bin4_NormVal,
         NL5reads_corrected = NL5reads / NL_bin5_NormVal,
         NL6reads_corrected = NL6reads / NL_bin6_NormVal,
         NL7reads_corrected = NL7reads / NL_bin7_NormVal,
         NL8reads_corrected = NL8reads / NL_bin8_NormVal,
         NL9reads_corrected = NL9reads / NL_bin9_NormVal,
         NL10reads_corrected = NL10reads / NL_bin10_NormVal,
         NL11reads_corrected = NL11reads / NL_bin11_NormVal,
         NL12reads_corrected = NL12reads / NL_bin12_NormVal
  ) 
```
## Compute a relative abundances matrix  
Biswas et al., 2021: Compute a relative abundance table, R, by dividing the columns of C by their sums. The columns of R sum to 1.  
```{r}

NL1_total <- sum(NL_allBC$NL1reads_corrected)
NL2_total <- sum(NL_allBC$NL2reads_corrected)
NL3_total <- sum(NL_allBC$NL3reads_corrected)
NL4_total <- sum(NL_allBC$NL4reads_corrected)
NL5_total <- sum(NL_allBC$NL5reads_corrected)
NL6_total <- sum(NL_allBC$NL6reads_corrected)
NL7_total <- sum(NL_allBC$NL7reads_corrected)
NL8_total <- sum(NL_allBC$NL8reads_corrected)
NL9_total <- sum(NL_allBC$NL9reads_corrected)
NL10_total <- sum(NL_allBC$NL10reads_corrected)
NL11_total <- sum(NL_allBC$NL11reads_corrected)
NL12_total <- sum(NL_allBC$NL12reads_corrected)

NL_allBC_R <- NL_allBC %>%
  mutate(NL1_norm=NL1reads_corrected/NL1_total,
         NL2_norm=NL2reads_corrected/NL2_total,
         NL3_norm=NL3reads_corrected/NL3_total,
         NL4_norm=NL4reads_corrected/NL4_total,
         NL5_norm=NL5reads_corrected/NL5_total,
         NL6_norm=NL6reads_corrected/NL6_total,
         NL7_norm=NL7reads_corrected/NL7_total,
         NL8_norm=NL8reads_corrected/NL8_total,
         NL9_norm=NL9reads_corrected/NL9_total,
         NL10_norm=NL10reads_corrected/NL10_total,
         NL11_norm=NL11reads_corrected/NL11_total,
         NL12_norm=NL12reads_corrected/NL12_total) %>%
  select(BC, NL1_norm, NL2_norm, NL3_norm, NL4_norm, NL5_norm, NL6_norm, NL7_norm, NL8_norm, NL9_norm, NL10_norm, NL11_norm, NL12_norm) %>%
  dplyr::rename(barcode=BC)

# Check, sum of all values in each column should equal 1
sum(NL_allBC_R$NL10_norm)
```
Calculate the total reads for each barcode across all bins  
```{r}
NL_allBC_total_counts <- NL_allBC %>% 
  mutate(BC_SorTotReads = NL1reads_corrected + NL2reads_corrected + NL3reads_corrected + NL4reads_corrected + NL5reads_corrected + NL6reads_corrected + NL7reads_corrected + NL8reads_corrected + NL9reads_corrected + NL10reads_corrected + NL11reads_corrected + NL12reads_corrected) %>% 
  select(BC, BC_SorTotReads) %>% 
  dplyr::rename(barcode=BC)
```
## Merge barcode-relative abundances with nucleotide sequences  
Read in barcode-nucleotide sequence mapping file. 
```{r}
consensus_gene <- read.csv(file="./input_files/consensus_gene.csv",head=TRUE,sep=",")
consensus_gene %>% select(description) %>% distinct() %>% nrow()  # 951065 unique barcodes

consensus_gene2 <- consensus_gene %>%
  select(description,sequence) %>%
  dplyr::rename(barcode=description,NTseq=sequence)

# convert to strings
str(consensus_gene2)
```
Merge from all bins and create a master list of unique variants and BCs
```{r}
tgood_NL <- right_join(consensus_gene2, NL_allBC_R, join_by(barcode)) %>%
  mutate(NTlen=nchar(as.character(NTseq)))

sum(is.na(tgood_NL$NTseq)) # the outputted number of barcodes in NL_allBC_R don't have a nucleotide sequence associated with them

# Normalization check, sum of all values in each column should still be 1
sum(tgood_NL$NL11_norm) 
```
Mapping check: how many barcodes have more than 1 variant?
```{r}
consensus_gene_sum <- consensus_gene2 %>%
  group_by(barcode) %>%
  summarise(count=n())

consensus_gene_sum %>%
  filter(count>1) %>%
  nrow(.) 
```
Make a list of BCs with only 1 variant and filter the dataset to keep only barcodes which have been mapped. 
```{r}
bcgood_NL <- consensus_gene_sum %>%
  filter(count==1) %>%
  select(-count)

# Filter to only keep barcodes which appear
NL_allBC_R_NTfilter <- tgood_NL %>%
  semi_join(bcgood_NL,by="barcode") %>% 
  left_join(NL_allBC_total_counts,by="barcode")
```
## Compute total normalized barcode counts for each bin
```{r}
NL_allBC_R_NTfilter_totals <- NL_allBC_R_NTfilter %>%
  group_by(barcode) %>%
  summarise(NTseq=NTseq,
            NL1_t=sum(NL1_norm),
            NL2_t=sum(NL2_norm),
            NL3_t=sum(NL3_norm),
            NL4_t=sum(NL4_norm),
            NL5_t=sum(NL5_norm),
            NL6_t=sum(NL6_norm),
            NL7_t=sum(NL7_norm),
            NL8_t=sum(NL8_norm),
            NL9_t=sum(NL9_norm),
            NL10_t=sum(NL10_norm),
            NL11_t=sum(NL11_norm),
            NL12_t=sum(NL12_norm),
            BC_SorTotReads=BC_SorTotReads)
```
## Merge barcode-relative abundances with amino acid sequences  
Load file of synTCS-MutLib variants - amino acid sequences - and filter to only keep translated sequences
```{r}
consensus_prot <- read.csv(file="./input_files/consensus_prot_with_PreSortBC.csv",head=TRUE,sep=",")

NL_allBC_R_AAfilter <- NL_allBC_R_NTfilter_totals %>%
  left_join(consensus_prot %>% dplyr::rename(barcode=BC),by="barcode")

# Some sequences have mutations which place a stop codon at the beginning and some barcodes were not mapped to amino acid sequences; filter these out.
NL_allBC_R_AAfilter <- NL_allBC_R_AAfilter %>%
  filter(!is.na(seq))
```
Replace read counts of NA and 0 with an arbitrary value of 0.1 for presort1 and presort2 libraries
```{r}
NL_allBC_R_AAfilter$presort1reads[is.na(NL_allBC_R_AAfilter$presort1reads)] <- 0.1
NL_allBC_R_AAfilter$presort1reads[NL_allBC_R_AAfilter$presort1reads == 0] <- 0.1
NL_allBC_R_AAfilter$presort2reads[is.na(NL_allBC_R_AAfilter$presort2reads)] <- 0.1
NL_allBC_R_AAfilter$presort2reads[NL_allBC_R_AAfilter$presort2reads == 0] <- 0.1
```
## Calculate a fold-change matrix  
Biswas et al., 2021: Divide each column of R element-wise by the input relative abundance vector (relative abundance of variants in the library before flow cytometry) to obtain a fold-change table, F.  
```{r}
NL_allBC_F <- NL_allBC_R_AAfilter %>%
  mutate(NL1_fc=NL1_t/presort2_norm,
         NL2_fc=NL2_t/presort2_norm,
         NL3_fc=NL3_t/presort2_norm,
         NL4_fc=NL4_t/presort2_norm,
         NL5_fc=NL5_t/presort2_norm,
         NL6_fc=NL6_t/presort2_norm,
         NL7_fc=NL7_t/presort2_norm,
         NL8_fc=NL8_t/presort2_norm,
         NL9_fc=NL9_t/presort2_norm,
         NL10_fc=NL10_t/presort2_norm,
         NL11_fc=NL11_t/presort2_norm,
         NL12_fc=NL12_t/presort2_norm)
```
## Calculate an adjusted abundances table  
Biswas et al., 2021: Divide each row of F by its sum to obtain a table of adjusted abundances, A. Each row of A sums to 1.  
```{r}
NL_allBC_A <- NL_allBC_F %>%
  mutate(rowsum=NL1_fc+NL2_fc+NL3_fc+NL4_fc+NL5_fc+NL6_fc+NL7_fc+NL8_fc+NL9_fc+NL10_fc+NL11_fc+NL12_fc) %>%
  mutate(NL1=NL1_fc/rowsum,
         NL2=NL2_fc/rowsum,
         NL3=NL3_fc/rowsum,
         NL4=NL4_fc/rowsum,
         NL5=NL5_fc/rowsum,
         NL6=NL6_fc/rowsum,
         NL7=NL7_fc/rowsum,
         NL8=NL8_fc/rowsum,
         NL9=NL9_fc/rowsum,
         NL10=NL10_fc/rowsum,
         NL11=NL11_fc/rowsum,
         NL12=NL12_fc/rowsum)
```
## Determine median bin for barcodes via cumulative sum across adjusted abundances
```{r}
# create functions to compute upper and lower index
maxcs = function(x, output){
  return(max(which(c(x[1],x[2],x[3],x[4],x[5],x[6],x[7],x[8],x[9],x[10],x[11],x[12]) < 0.5)))
}
mincs = function(x, output){
  return(min(which(c(x[1],x[2],x[3],x[4],x[5],x[6],x[7],x[8],x[9],x[10],x[11],x[12]) >= 0.5)))
}

# Compute cumulative sum across adjusted barcode abundances for all bins to estimate median bin
NL_allBC_CS <- NL_allBC_A %>%
  rowwise() %>%
  mutate(
    cumulative_p = list(cumsum(c(NL1,NL2,NL3,NL4,NL5,NL6,NL7,NL8,NL9,NL10,NL11,NL12))),
    lower_index = max(which(cumsum(c(NL1,NL2,NL3,NL4,NL5,NL6,NL7,NL8,NL9,NL10,NL11,NL12)) < 0.5)),
    upper_index = min(which(cumsum(c(NL1,NL2,NL3,NL4,NL5,NL6,NL7,NL8,NL9,NL10,NL11,NL12)) >= 0.5)),
    median = ifelse(
      is.infinite(lower_index), 
      1, 
      lower_index + (0.5 - unlist(cumulative_p)[lower_index]) / 
        (unlist(cumulative_p)[upper_index] - unlist(cumulative_p)[lower_index])
    )
  )
```
## Export the data as a CSV file
```{r}
NL_allBC_CS_toCSV <- NL_allBC_CS %>%
  select(barcode, NTseq, seq, lower_index, upper_index, median, BC_SorTotReads, presort1reads, presort2reads)

write.csv(NL_allBC_CS_toCSV,
          "./output_files/DcuS_NoLigand_bin_distribution-byBC.csv", row.names = FALSE)

```


# Reproducibility

The session information is provided for full reproducibility.
```{r}
devtools::session_info()
```

# References
Biswas, S.; Khimulya, G.; Alley, E. C.; Esvelt, K. M.; Church, G. M. Low-N Protein Engineering with Data-Efficient Deep Learning. Nat. Methods 2021, 18 (4), 389–396. https://doi.org/10.1038/s41592-021-01100-y. 