---
title: "Characterizing Sequence-Function Relationships in Chimeric DcuS/EnvZ Histidine Kinases"
author: 'Authors: <font color="blue">Luca B. Lippert, </font color="black">Samuel R. Hinton, Andrew Holston, Karl J. Romanowicz, Calin Plesa'
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
      smooth_scroll: yes
  pdf_document:
    toc: yes
    toc_depth: '5'
---

<font color="black"> **R Notebook:** <font color="green">Provides reproducible analysis for **Statistical analysis of multi-barcoded mutant sequences** in the following manuscript: <font color="black">

**Citation:** Lippert LB, Hinton SR, Holston A, Romanowicz KJ, Plesa C. Characterizing Sequence-Function Relationships in Chimeric DcuS/EnvZ Histidine Kinases. *In Prep.* 2026.

**GitHub Repository:** <https://github.com/PlesaLab/DcuSEnvZ>

# Experiment

<font color="black">This pipeline processes</font> <font color="red">*barcode-sequence-phenotype data previously generated in Merge_and_MEFL_Convert.Rmd* <font color="black"> and determines which multi-barcoded mutnts have significant differences between their fold-change scores.

```{css, echo=FALSE}
.badCode {
background-color: lightpink;
font-weight: bold;
}

.goodCode {
background-color: lightgreen;
font-weight: bold;
}

table {
  margin: auto;
  border-top: 1px solid #666;
  border-bottom: 1px solid #666;
}
table thead th { border-bottom: 1px solid #ddd; }
th, td { padding: 5px; }
thead, tfoot, tr:nth-child(even) { background: #eee; }
```

```{r setup, include=FALSE, echo=FALSE}
# Set global options for notebook
knitr::opts_chunk$set(
  echo = TRUE,
  warning = TRUE,
  message = TRUE,
  results = "markup",
  collapse = FALSE
)
# Getting the path of your current open file and set as wd
current_path = rstudioapi::getActiveDocumentContext()$path 
setwd(dirname(current_path))
print(getwd())
```

# Packages

The following R packages must be installed prior to loading into the R session. See the **Reproducibility** tab for a complete list of packages and their versions used in this workflow.

```{r message=FALSE, results='hide', warning=FALSE}
# Make a vector of required packages
required.packages <- c("devtools", "knitr", "patchwork", "tidyverse", "ggplot2", "dplyr", "tidyr", "magrittr", "stringr", "seqinr", "purrr", "FSA", "rstatix")

# Load required packages
lapply(required.packages, library, character.only = TRUE)
```

```{r include=FALSE}
# set.seed is used to fix the random number generation to make the results repeatable
set.seed(123)
```

# Dial-Out Mutant Selection Pipeline

<font color="blue">**This section is based on the R file: "Dial_Out_Mutant_Selection.R".**<font color="black"> It describes how to load all of the pre-existing barcode-sequence-phenotype data necessary for downstream analysis. Here, we use the Kruskal-Walllis (a nonparametric ANOVA) and post-hoc Dunn statistical tests to determine if there are significant differences between the fold-change scores observed for all multi-barcoded mutants. These mutants determined to have significant differences between their phenotype scores were selected from for dial-out PCR and individual characterization. The end result of this script is two .CSV files containing a set of mutant amino acid sequences, their associated barcodes, and phenotype scores.

## Read in data

```{r}
merged_BC_MEFL_norm <- read.csv(file="./output_files/All_DcuS_Data-medianMEFL.csv", head=T, sep=",")
```

## Wrangle and filter the data

...the data to associate phenotype with the specific amino acid mutation found in mutant sequences.

For each mutant DcuS sequence, align it to wild-type sequence and determine which amino acids differ. Record three columns: amino acid, position, mutation

```{r}
# Filter to keep only mutants that are the correct length (192 amino acids) and are associated with more than one barcode
multiBC_data <- merged_BC_MEFL_norm %>%
  filter(str_length(AAseq)==192) %>%
  group_by(AAseq) %>%
  filter(n_distinct(barcode) > 1) %>%
  ungroup()

# make a vector of the wild-type sequence to compare mutant sequences to
wt_seq = "RHSLPYRMLRKRPMKLSTTVILMVSAVLFSVLLVVHLIYFSQISDMTRDGLANKALAVARTLADSPEIRQGLQKKPQESGIQAIAEAVRKRNDLLFIVVTDMQSLRYSHPEAQRIGQPFKGDDILKALNGEENVAINRGFLAQALRVFTPIYDENHKQIGVVAIGLELSRVTQQINDSRWSLQMAAGVKQLA"

# WT sequence as a vector
wt_seq_vec <- str_split(wt_seq, "")[[1]]

# custom mutation parser
find_mutations <- function(mut_seq) {
  mut_vec <- str_split(mut_seq, "")[[1]]
  # safety check
  if (length(mut_vec) != length(wt_seq_vec)) {
    return(NA_character_)
  }
  diff_pos <- which(mut_vec != wt_seq_vec)
  # no mutations
  if (length(diff_pos) == 0) {
    return(NA_character_)
  }
  # output of the function:
  paste0(
    wt_seq_vec[diff_pos],
    diff_pos +1,
    mut_vec[diff_pos],
    collapse = "_"
  )
}

multiBC_data <- multiBC_data %>% 
  # Create the mutation identifier string
  mutate(mutations = sapply(AAseq, find_mutations)) 
```

## Statistical test on fold-change scores

Define a function to perform the Kruskal-Wallis and Dunn tests for each observed multi-barcoded mutant

```{r}
analyze_mutant_wide_FoldChange <- function(df_mutant) {
  # number of barcodes for each mutant
  n_barcodes = nrow(df_mutant)
  seq = df_mutant$AAseq
  
  # Pivot only inside the function for stats
  df_long <- df_mutant %>%
    pivot_longer(
      cols = c(NoLig_fc, Fum_fc, Asp_fc),
      names_to = "Condition",
      values_to = "MEFL"
    )
  # Kruskal–Wallis
  kw <- kruskal.test(MEFL ~ Condition, data = df_long)
  # Dunn post-hoc (safe)
  dunn <- tryCatch({
    dunnTest(MEFL ~ Condition, data = df_long, method = "bh")$res
  }, error = function(e) NULL)
  # Medians computed directly from wide-format columns
  
  med_NoLig <- median(df_mutant$NoLig_fc, na.rm = TRUE)
  med_Fum   <- median(df_mutant$Fum_fc, na.rm = TRUE)
  med_Asp   <- median(df_mutant$Asp_fc, na.rm = TRUE)
  # Create a tibble of medians and deltas
  medians <- tibble(
    Condition = c("NoLig_fc_med", "Fum_fc_med", "Asp_fc_med"),
    median_MEFL = c(med_NoLig, med_Fum, med_Asp),
  )
  # Return everything
  list(
    seq = seq,
    kw_p = kw$p.value,
    dunn = dunn,
    Medians = medians,
    n_barcodes = n_barcodes
  )
}
```

Run the statistical tests directly from multiBC_data

```{r, warning=FALSE}
results_FoldChange <- multiBC_data %>%
  group_by(AAseq) %>%
  group_map(~ analyze_mutant_wide_FoldChange(.x), .keep = TRUE)

names(results_FoldChange) <- map_chr(results_FoldChange, ~ unique(.x$seq))
```

Create a summary table of the statistical test results

```{r}
summary_table_FoldChange <- map_df(
  names(results_FoldChange),
  function(m) {
    res <- results_FoldChange[[m]]
    # Extract barcode lists and values for this mutant
    mutant_df <- multiBC_data %>% filter(AAseq == m)
    barcodes <- mutant_df$barcode %>% unique() %>% paste(collapse = ",")
    
    NoLig_Median_MEFL_median = median(mutant_df$NoLig_Median_MEFL)
    Fum_Median_MEFL_median = median(mutant_df$Fum_Median_MEFL)
    Asp_Median_MEFL_median = median(mutant_df$Asp_Median_MEFL)
    
    NoLig_fc_median = median(mutant_df$NoLig_fc)
    Fum_fc_median = median(mutant_df$Fum_fc)
    Asp_fc_median = median(mutant_df$Asp_fc)
    
    Fum_DNR_median = median(mutant_df$Fum_DNR)
    Asp_DNR_median = median(mutant_df$Asp_DNR)
    Lig_spec_median = median(mutant_df$Lig_spec)
    
    # Handle Dunn p-values safely
    if (!is.null(res$dunn) && "P.adj" %in% colnames(res$dunn)) {
      dmin <- suppressWarnings(min(res$dunn$P.adj, na.rm = TRUE))
      if (is.infinite(dmin)) dmin <- NA
    } else {
      dmin <- NA
    }
    # Handle effect size safely
    if (!is.null(res$Medians) && "delta_vs_NoLig" %in% colnames(res$Medians)) {
      deltas <- suppressWarnings(as.numeric(res$Medians$delta_vs_NoLig))
      max_delta <- if (all(is.na(deltas))) NA else max(abs(deltas), na.rm = TRUE)
    } else {
      max_delta <- NA
    }
    tibble(
      seq = m,
      barcodes = barcodes,
      n_barcodes = str_count(barcodes, ",") + 1,
      
      KW_p = res$kw_p,
      Dunn_min_padj = dmin,
      Max_abs_delta = max_delta,
  
      NoLig_Median_MEFL_median = NoLig_Median_MEFL_median,
      Fum_Median_MEFL_median = Fum_Median_MEFL_median,
      Asp_Median_MEFL_median = Asp_Median_MEFL_median,
      
      NoLig_fc_median = NoLig_fc_median,
      Fum_fc_median = Fum_fc_median,
      Asp_fc_median = Asp_fc_median,
      
      Fum_DNR_median = Fum_DNR_median,
      Asp_DNR_median = Asp_DNR_median,
      Lig_spec_median = Lig_spec_median,

    )
  }
)
```

Filter to keep only mutants with a Kruskal-Wallis p-value less than 0.05

```{r}
significant_mutants_FoldChange <- summary_table_FoldChange %>%
  filter(KW_p < 0.05)
```

Export the dial-out mutant data

```{r}
write.csv(significant_mutants_FoldChange, "./output_files/significant_mutants_FoldChange.csv", row.names = FALSE)
```

## Statistical test on inferred MEFL values

Define a function to perform the Kruskal-Wallis and Dunn tests for each observed multi-barcoded mutant

```{r}
analyze_mutant_wide_MedianMEFL <- function(df_mutant) {
  # number of barcodes for each mutant
  n_barcodes = nrow(df_mutant)
  seq = df_mutant$AAseq
  
  # Pivot only inside the function for stats
  df_long <- df_mutant %>%
    pivot_longer(
      cols = c(NoLig_Median_MEFL, Fum_Median_MEFL, Asp_Median_MEFL),
      names_to = "Condition",
      values_to = "MEFL"
    )
  # Kruskal–Wallis
  kw <- kruskal.test(MEFL ~ Condition, data = df_long)
  # Dunn post-hoc (safe)
  dunn <- tryCatch({
    dunnTest(MEFL ~ Condition, data = df_long, method = "bh")$res
  }, error = function(e) NULL)
  # Medians computed directly from wide-format columns
  
  med_NoLig <- median(df_mutant$NoLig_Median_MEFL, na.rm = TRUE)
  med_Fum   <- median(df_mutant$Fum_Median_MEFL, na.rm = TRUE)
  med_Asp   <- median(df_mutant$Asp_Median_MEFL, na.rm = TRUE)
  # Create a tibble of medians and deltas
  medians <- tibble(
    Condition = c("NoLig_Median_MEFL_med", "Fum_Median_MEFL_med", "Asp_Median_MEFL_med"),
    median_MEFL = c(med_NoLig, med_Fum, med_Asp),
  )
  # Return everything
  list(
    seq = seq,
    kw_p = kw$p.value,
    dunn = dunn,
    Medians = medians,
    n_barcodes = n_barcodes
  )
}
```

Run the statistical tests directly from multiBC_data

```{r, warning=FALSE}
results_MedianMEFL <- multiBC_data %>%
  group_by(AAseq) %>%
  group_map(~ analyze_mutant_wide_MedianMEFL(.x), .keep = TRUE)

names(results_MedianMEFL) <- map_chr(results_MedianMEFL, ~ unique(.x$seq))
```

Create a summary table of the statistical test results

```{r}
summary_table_MedianMEFL <- map_df(
  names(results_MedianMEFL),
  function(m) {
    res <- results_MedianMEFL[[m]]
    # Extract barcode lists and values for this mutant
    mutant_df <- multiBC_data %>% filter(AAseq == m)
    barcodes <- mutant_df$barcode %>% unique() %>% paste(collapse = ",")
    
    NoLig_Median_MEFL_median = median(mutant_df$NoLig_Median_MEFL)
    Fum_Median_MEFL_median = median(mutant_df$Fum_Median_MEFL)
    Asp_Median_MEFL_median = median(mutant_df$Asp_Median_MEFL)
    
    NoLig_fc_median = median(mutant_df$NoLig_fc)
    Fum_fc_median = median(mutant_df$Fum_fc)
    Asp_fc_median = median(mutant_df$Asp_fc)
    
    Fum_DNR_median = median(mutant_df$Fum_DNR)
    Asp_DNR_median = median(mutant_df$Asp_DNR)
    Lig_spec_median = median(mutant_df$Lig_spec)
    
    # Handle Dunn p-values safely
    if (!is.null(res$dunn) && "P.adj" %in% colnames(res$dunn)) {
      dmin <- suppressWarnings(min(res$dunn$P.adj, na.rm = TRUE))
      if (is.infinite(dmin)) dmin <- NA
    } else {
      dmin <- NA
    }
    # Handle effect size safely
    if (!is.null(res$Medians) && "delta_vs_NoLig" %in% colnames(res$Medians)) {
      deltas <- suppressWarnings(as.numeric(res$Medians$delta_vs_NoLig))
      max_delta <- if (all(is.na(deltas))) NA else max(abs(deltas), na.rm = TRUE)
    } else {
      max_delta <- NA
    }
    tibble(
      seq = m,
      barcodes = barcodes,
      n_barcodes = str_count(barcodes, ",") + 1,
      
      KW_p = res$kw_p,
      Dunn_min_padj = dmin,
      Max_abs_delta = max_delta,
  
      NoLig_Median_MEFL_median = NoLig_Median_MEFL_median,
      Fum_Median_MEFL_median = Fum_Median_MEFL_median,
      Asp_Median_MEFL_median = Asp_Median_MEFL_median,
      
      NoLig_fc_median = NoLig_fc_median,
      Fum_fc_median = Fum_fc_median,
      Asp_fc_median = Asp_fc_median,
      
      Fum_DNR_median = Fum_DNR_median,
      Asp_DNR_median = Asp_DNR_median,
      Lig_spec_median = Lig_spec_median,

    )
  }
)
```

Filter to keep only mutants with a Kruskal-Wallis p-value less than 0.05

```{r}
significant_mutants_MedianMEFL <- summary_table_MedianMEFL %>%
  filter(KW_p < 0.05)
```

Export the dial-out mutant data

```{r}
write.csv(significant_mutants_MedianMEFL, "./output_files/significant_mutants_MedianMEFL.csv", row.names = FALSE)
```

# Reproducibility

The session information is provided for full reproducibility.

```{r}
devtools::session_info()
```
